---
title: "Masters Homework 2"
author: "Pranavasri Rudrakshala Matam"
date: "7/21/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Credit received ## 

>- Find summary information about the total blood mercury levels in our NYC HANES data (hint: names(nyc) will call the names of each column). What is the max value, what is the minimum value.  

max value = 129.2344
min value = 0.0919

```{r}
# names(nyc)

nyc_summary <- summary(nyc$MERCURYTOTALBLOOD)
nyc_summary
```

>- Bonus/Challenge: What is the Standard Deviation and skewness of the blood mercury?

```{r}
# Standard Deviation 
sd_total_mercury <- sd(nyc$MERCURYTOTALBLOOD, na.rm = TRUE)
print(sd_total_mercury)

# Skewness 
data <- nyc$MERCURYTOTALBLOOD
data <- data[!is.na(data)]  # Remove NA values

n <- length(data)
mean_data <- mean(data)
sd_data <- sd(data)

skewness_total_mercury <- (n * sum((data - mean_data)^3)) / ((n - 1) * (n - 2) * sd_data^3)
print(skewness_total_mercury)
```

>- Install a package and load, such as "ggplot2" or "dplyr"

```{r}
install.packages("ggplot2")
library(ggplot2)
```


>- What two commands are needed to install and use a package in R?

Two commands needed to install and use packages are: install.packages("package_name) and library("package_name").

```{r}
install.packages("dplyr")
library(dplyr)
```


>- When you read in a .csv file into R, what class is the object that you read in (like nyc)?

The class of the object that you read in is a data frame.


>- What command can you use to determine what type of data is in a column?

Both the commands class() and typeof() can provide the type of data in the column. Class() is the more commonly used method and tells us what the object is whereas typeof() tells us what the object is made of at an internal level.

```{r}
class(nyc$AGEGROUP)
class(nyc$UALBUMIN)

typeof(nyc$AGEGROUP)
typeof(nyc$UALBUMIN)
```

>- Provide a command to find the type of data in the DX_DBTS column of nyc. 

```{r}
class(nyc$DX_DBTS)
```

>- Provide the same command the type of data but use the other way of indexing from what you did above. 

```{r}
class(nyc[["DX_DBTS"]])
```

>- What would happen if we had chosen the first 4 columns of NYC HANES and made a matrix? What would the class of the cells in this matrix be? Why?

When you select 4 columns of NYC HANES and convert them to a matrix, the class of the cells will be 'character' as R converts all the values to the same type in a matrix to maintain structure.

```{r}
subset_nyc <- nyc[, 1:4]
nyc_matrix <- as.matrix(subset_nyc)

class(nyc_matrix)
# head(nyc_matrix)
class(nyc_matrix[, 1])
class(nyc_matrix[, 2])
class(nyc_matrix[, 3])
class(nyc_matrix[, 4])

```

>- If we make a vector,vec<- c(3,9,4,7,8,9), and multiply it by WannaMatrix, what is the value in the third row, third column? 

The value of third row, third column is 4.

```{r}
WannaMatrix <- nyc[5:10, 6:9]
WannaMatrix <- as.matrix(WannaMatrix,nrow = 6, ncol = 4)

vec <- c(3, 9, 4, 7, 8, 9)

result <- WannaMatrix * vec

print(result[3, 3])
```


>- What happens if instead we made a vector vec2 <- c(3,9,4,7,8) and multiply it by WannaMatrix, what is the third row, third column value? What is going on? 

When we multiply wannaMatrix and vec2, we get a warning that says "longer object length is not a multiple of shorter object length". This is because the matrix has 6 rows but the vector only has 5. Despite the warning, R still performs the operation by repeating the vector: c(3, 9, 4, 7, 8, 3). Therefore we get the value of third row, third column as '8'.  

```{r}
vec2 <- c(3, 9, 4, 7, 8)

result2 <- WannaMatrix * vec2

print(result2[3, 3])
```


>- Look up the arguments for read.csv

```{r}
?read.csv
```


>- Use apply() to find the median value of CADMIUM, LEAD CHOLESTEROLTOTAL, GLUCOSESI, CREATININESI, and CREATININE

```{r}
apply(nyc[, c("CADMIUM", "LEAD", "CHOLESTEROLTOTAL", "GLUCOSESI", "CREATININESI", "CREATININE")], 2, median, na.rm = TRUE)

```


>- Do the same using sapply()

```{r}
sapply(nyc[, c("CADMIUM", "LEAD", "CHOLESTEROLTOTAL", "GLUCOSESI", "CREATININESI", "CREATININE")], median, na.rm = TRUE)
```

>- Use a for loop to find the median value of columns 16:19

```{r}
for (i in 16:19) {
  col_name <- colnames(nyc)[i]
  med <- median(nyc[[i]], na.rm = TRUE)
  cat("Median of", col_name, ":", med, "\n")
}
```


>- Make a for loop to find the standard deviation (hint sd() ) of CHOLESTEROLTOTAL, GLUCOSESI, CREATININESI, AND CREATININE

```{r}
cols <- c("CHOLESTEROLTOTAL", "GLUCOSESI", "CREATININESI", "CREATININE")

for (col in cols) {
  std_dev <- sd(nyc[[col]], na.rm = TRUE)
  cat("Standard deviation of", col, ":", std_dev, "\n")
}

```


>- Bonus/Challenge Create a vector in which to put the standard deviations 

```{r}
cols <- c("CHOLESTEROLTOTAL", "GLUCOSESI", "CREATININESI", "CREATININE")

# An empty numeric vector
std_devs <- numeric(length(cols))  

for (i in seq_along(cols)) {
  std_devs[i] <- sd(nyc[[cols[i]]], na.rm = TRUE)
}

names(std_devs) <- cols

print(std_devs)

```


>- Use ifelse to create a new variable of high and normal A1C dichotomizing with 7 as a threshold for high levels

```{r}
nyc$A1C_level <- ifelse(nyc$A1C >= 7, "High", "Normal")

table(nyc$A1C_level)
```

>- Subset to only have females (gender == "female")

```{r}
# Assuming "2" under gender in nyc df refers to females

# nyc_females <- subset(nyc, GENDER == "female")

nyc_females <- subset(nyc, GENDER == 2)
print(nyc_females)
```


>- Create a subset of males and find the median total cholesterol level

```{r}

# Assuming "1" under gender in nyc df refers to male

# nyc_male <- subset(nyc, GENDER == "male")

nyc_male <- subset(nyc, GENDER == "1")

# Median of CHOLESTEROLTOTAL
median_cholesterol_male <- median(nyc_male$CHOLESTEROLTOTAL, na.rm = TRUE)

print(median_cholesterol_male)

```


>- Try to create a summary of the mean A1C using two factors, GENDER and DX_DBTS

```{r}

# tapply(nyc$A1C, list(nyc$GENDER, nyc$DX_DBTS), mean, na.rm = TRUE)

aggregate(A1C ~ GENDER + DX_DBTS, data = nyc, FUN = mean, na.rm = TRUE)


```

Find the class of the first three columns of data. What are they?

The class of the first three columns are: "character", "integer" and "integer".

```{r}
sapply(nyc[ , 1:3], class)

```


